The Sub-Millimetre User Reduction Facility: an iterative map-maker for \scuba

### 2012-08-23 smurf_paper.tex
comments from Per Friberg
These were actually from early June I think. I screwed up a
merge at some point, and had to add these in again later. Some
points are not addressed yet, so I left "Per: bla bla" comments
in a few places as a reminder

The Submillimetre Common User Bolometer Array 2
\citep[\scuba,][]{2006SPIE.6275E..45H,holland2012} is a new instrument
for the 15-m James Clerk Maxwell Telescope (JCMT) on Mauna Kea,
Hawai'i. The camera simultaneously images the sky in two broad bands
centered over 450 and 850\,\micron, with approximately 7.5 and
14.5\,arcsec full-width at half-maximum (FWHM) point spread functions
(PSFs). The focal planes at each wavelength are populated with 4
rectangular subarrays, consisting of $40 \times 32$ bolometers each,
and subtend a nearly contiguous 7\,arcmin $\times$ 7\,arcmin footprint
on the sky.  The instrument is also equipped with a Fourier transform
spectrometer \citep[FTS-2,][]{gom2010}, and a polarimeter
\cite[POL-2,][]{2011ASPC..449...68B}. This paper describes the
properties of \scuba\ data that are relevant to producing maps of the
imaging data, and the Submillimetre User Reduction Facility, SMURF, a
software package for perfoming the reduction written using the
Starlink Software Environment \citep{2009ASPC..411..418J}. The
analysis of FTS-2 and POL-2 data will be described at a later date
once these additional instruments have been commissioned and
characterized. The details of the instrument design, performance, and
its calibration are given in two companion papers: \citet{holland2012}
and \citet{dempsey2012}.

@@

The \scuba\ bolometer temperatures are read-out using superconducting
transition-edge sensors (TESs), with critical temperatures
$T_\mathrm{C}$ (...typical value...). The range about $T_\mathrm{C}$
in which the transition from a superconductor to a normal material
occurs is extremely small (...typical value...?). If the devices are
biased to a temperature within this transition on large time scales,
then tiny temperature fluctuations on short time scales will produce
significant changes in the impedence. The entire \scuba\ focal plane
is therefore held at a \emph{base temperature} somewhat below
$T_\mathrm{C}$. The TESs then have small electrical voltage biases,
and individual pixel heaters (using a current bias to apply the
desired Joule heating) are used to place the detectors in the
transition. Only a small detector bias voltage is required since the
heaters compensate for most of the variations in the base temperature,
and other ambient loads (like the atmosphere).  Small temperature (and
therefore impedence) fluctuations produce a changing current through
the TESs. The resulting changing magnetic fields are then amplified
using chains of superconducting quantum interference devices (SQUIDs),
and the larger output current is finally digitized. Each detector has
its own SQUID for the first-stage of the amplification, but the
remaining stages occur within a common chain of SQUIDs for each column
of 40 detectors. Each of the 32 rows is then read-out in sequence at a
row-visit frequency of ...kHz?... This multiplexing scheme greatly
reduces the number of readout and bias lines to each detector, making
such large-format bolometer array designs feasible. It should also be
noted that there is an additional 33rd row of SQUID readouts that are
not connected to TES. These ``dark SQUIDs'' are used to track
non-thermal noise sources that are common to each column's amplifier
chain.

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. Since there
are variations in $T_\mathrm{C}$ across each subarray due to
manufacturing differences, considerable effort has gone into
dynamically applying position-dependent biases as well as a total
focal plane bias to make as many detectors usable as possible
\citep{holland2012}. In practice, the response of a given detector can
therefore vary somewhat from observation to observation, depending on
how the bias and amplification parameters were set up. To track these
responsivity changes, at the start of each observation, the current is
ramped through the bolometer heaters such that $dI/dP$ can be
measured. Note that the bolometers may have a slightly different
response to heater power due to conductance as compared to the
response to optical power, so additioanl efficiency corrections must
also be obtained. While not discussed further in this paper, details
may be found in \citet{dempsey2012}. We will refer to measurements of
$dI/dP$ across the focal plane as the
\emph{flatfield}. Fig.~\ref{fig:sensitivities} shows the flatfield for
each focal plane.

%\begin{figure}
%\vspace{2cm}
%\textbf{This will be a figure showing s4a and s8d responsivity maps}
%\vspace{2cm}
%\caption{The response of each bolometer to known input heater power
%  for a typical night during S2SRO. The s4a subarray had a
%  particularly large variation in $T_\mathrm{C}$ towards the
%  bottom-right which resulted in lower responsivity, and
%  correspondingly larger noise. The s8d array also suffered
%  significant variations across the focal plane, in particular
%  removing larger number of useful detectors from the bottom-left.}
%\label{fig:sensitivities}
%\end{figure}

In general, bolometers have two predominant noise components: (i) the
effectively white thermal noise produced by the phonon and optical
loading on the absorbers; and (ii) and longer timescale drift
resulting from slow variations in the mean load (e.g. fridge,
telescope, or atmospheric changes), and also the readout
electronics. As such, the spectrum corresponding to a bolometer
time-series exhibits a ``$1/f$'' shape below some knee frequency,
$f_\mathrm{knee}$, and is relatively flat at higher frequencies. Other
smaller effects may include line features in the spectrum (a common
culprit is pickup from AC mains, or microphonic signals induced by
mechanical vibrations), and finally a roll-off towards the Nyquist
frequency caused by the anti-aliasing filter that precedes
digitization.

The final conversion of power received by the detector into
astronomical flux units requires two more fundamental corrections: (i)
atmospheric extinction (due to atmospheric water vapour) must be
compensated for as a function of time; and (ii) the corrected power is
related to an astronomical flux density through the effective area of
the detector, the spectral response of the bolometers (bandpass), the
spectrum of the source, and additional optical efficiency factors -- a
\emph{flux conversion factor} (FCF). Extinction is measured, in most
cases, using the JCMT line-of-sight water vapour radiometer (WVM)
which is calibrated using sky-dips and the CSO 225\,GHz tipper. The
FCF is determined empirically using measurements of standard
sources. For further details on these calibration steps, see
\citet{dempsey2012}.

@@

In order to observe astronomical sources, it is clearly advantageous
that the signals appear in the region of the spectrum that offers the
highest \snr, namely frequencies above $f_\mathrm{knee}$. This effect
is achieved by modulating the signal; the original SCUBA primarily
used the JCMT chopping secondary to place sources into the spectrum at
about $\sim$10\,Hz. The problem with this strategy was that SCUBA was
ultimately limited to recovering information on angular scales smaller
than the chop throw (a maximum of several arcmin). The primary
approach to map-making with \scuba\ is instead to slew the entire
telescope at such a rate that angular scales of interest are crossed
on time scales shorter than $1/f_\mathrm{knee}$, similar to other
ground-based instruments (MAMBO, BOLOCAM, SHARC-II, AzTEC,
LABOCA,...). The desire to scan rapidly has two basic constraints: (i)
the telescope cannot physically move faster that several hundred
arcsec\,sec$^{-1}$ without requiring unacceptably large accelerations
to change direction; and (ii) bolometers are designed with a thermal
time constant that balances response with sensitivity. Taking these
constraints into consideration, \scuba\ was ultimately designed to
work at a maximum scan speed of 600\,arcsec\,sec$^{-1}$. The data
acquisition system has a target sample rate of 200\,Hz, which at that
scan velocity results in a sample every 3.0\,arcsec, giving
approximately 2.5 samples per 450\,\micron\ diffraction-limited
full-width at half-maximum (FWHM) -- a typical rule-of-thumb for
adequately sampling a Gaussian point spread function (PSF). Similarly,
the \scuba\ bolometers are designed to have a time constant of 2 ms
\textbf{check that number!} in order to avoid distorting PSF-sized
sources at this maximum scan rate.

The scan strategy is designed, primarily, to mitigate the effects of
$1/f$ drift. The simplest scan one might use is a raster
``boustrophedon'' pattern, i.e. a series of parallel passes across the
field. However, such a scan yields extremely poor performace for
structures that are transverse to the scan direction, since a long
time will have ellapsed between each parallel step, and the signal
will therefore be totally dominated by the $1/f$ noise. In addition,
other periodic noise features may appear in the data. With such a
regularly spaced scan strategy, these noise features can easily
conspire with the scan pattern to produce similarly periodic noise
structures in the map, or \emph{scan-synchronous} noise. It has
therefore been recognized that scan strategies that visit each region
of the map on as many different time scales, and scan angles as
possible offer the best protection against large-scale noise
\citep[e.g.][]{tegmark1997,kovacs2008b}; such strategies are said to
offer good \emph{cross-linking}. As a tradeoff between truly random
scan patterns and feasibility of implementation taking into
consideration both mechanical and scheduling constraints, \scuba
presently has three primary scan modes: PONG (cite Borys et al.),
LISSAJOUS, and DAISY... (more citations? Kovacs et al. has a nice
description... also CMB literature).

@@

Next, we produce power spectral density (PSD) plots for several
representative bolometers in Fig.~\ref{fig:pspec}. To produce this
figure, we follow the convention that the PSD as a function of
frequency, $\mathbf{P}(f)$, is normalized such that the integral over
frequency gives the same variance as the time-series variance across
the full time series. In other words, given a bolometer signal
$\mathbf{b}(t)$,
%
\begin{equation}
\label{eq:psd}
\langle\mathbf{b}^2(t)\rangle = \int \mathbf{P}(f) df .
\end{equation}
%
However, we only show the PSD up to the Nyquist frequency, so the
conversion from variance measured in the frequency domain to the time
domain requires an additional multiplcation by 2; hence the units are
pW$^2$\,Hz$^{-1}$ rather than pW$^2$\,s.

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{pspec.pdf}
\caption{Bolometer power spectral densities. Four representative
  bolometers have been selected from each sub-array, and the
  flat-fielded (but otherwise raw) PSDs are shown as coloured dotted
  lines (the blue signals are for the same time series as those shown
  in Fig.~\ref{fig:bolos_mix}). Horizontal dotted lines at $5 \times
  10^{-8}$ and $5 \times 10^{-9}$\,pW\,Hz$^{-1}$ at 450 and
  850\,\micron, respectively, give the approximate white-noise
  levels. There are broad line features in the PSDs at both
  wavelengths above $\sim$35\,Hz. From $\sim$50\,Hz up to the Nyquist
  frequency, the gradual roll-off of the anti-aliasing filter is also
  evident.  At lower frequencies, the bolometer signals exhibit clear
  $1/f$ knees at approximately 2 and 4\,Hz at 450 and
  850\,\micron. The solid black lines are the PSDs of the common-mode
  signals at each wavelength, and the solid coloured lines show the
  PSDs of the bolometers once the common-mode is removed. These
  residual signals have significantly lower $1/f$ knees, approximately
  0.4 and 2\,Hz at 450 and 850\,\micron. Finally, the dashed black
  lines show the spectral shape produced by a point source, assuming
  that the telescope is slewing at 200\,arcsec\,sec$^{-1}$ (typical
  for S2SRO); for reference, the top horizontal axis shows the
  conversion from frequency to angular scale. We can see that the line
  features may add significant noise on the scale of the PSF. We also
  see that, for this scan velocity, the low-frequency noise begins to
  dominate on scales of approximately 8.3 and 1.7\,arcmin; these
  scales are comparable to the array footprint.}
\label{fig:pspec}
\end{figure*}

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded data. At each wavelength, there is a clear
$1/f$ knee at a few Hz, followed by a predominantly white spectrum
punctuated by line features above $\gsim 30$\,Hz, and finally a
roll-off caused by the anti-aliasing filter above $\gsim 70$\,Hz. As
indicated in the previous section, the correlation between the
low-frequency components of the different bolometer signals is
large. The solid black lines in Fig.~\ref{fig:pspec} indicate the PSDs
of the common-modes $\mathbf{c}(t)$ at each wavelength, which
reproduce most of the low-frequency structure, as well as the
high-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high-frequency,
as expected if the bolometers are dominated by un-correlated white
noise. Finally, we note that the amplitudes and slopes of
$\mathbf{c}(t)$ are similar at both 450 and 850\,\micron. However,
since the 450\,\micron\ white noise is larger than at 850\,\micron\
(shown approximately by the horizontal dotted lines), the $1/f$ knee
occurs at a \emph{lower} frequency at 450\,\micron.

@@

An example of an observation where pickup appears to be significant is
shown in Fig.~\ref{fig:magpickup}. The time-series for two
450\,\micron\ bolometers in the same column (not flatfielded) show
that there is a strong signal with a similar shape, but opposite
signs. This behaviour is seen across the entire array. The dark squid
signal for the same columns exhibits a similar shape and
amplitude. Since the dark squid has no thermal absorber or TES
attached to it, this observed signal is not likely to be optical or
thermal in nature (although there can be some cross-talk with the
bolometers). It is also for this reason that the higher-frequency
noise that appears in the bolometer data (thermal) does not appear in
the dark squid, leading to a significantly higher \snr.

@@

The telescope pointing offsets for this approximately 0.5\,degree
diameter scan are also shown in Fig.~\ref{fig:magpickup}. Since the
phase of the azimuth offsets from the map centre in this scan pattern
slowly drifts with respect to the elevation offsets, it is clear that
the bolometer and dark squid signals are detecting a noise source that
is correlated only with the azimuthal motion and \emph{not} the
elevation. This behaviour would be expected if if there were a strong
magnetic field fixed with respect to the telescope dome (i.e., the
earth's magnetic field). Since \scuba\ is mounted on a Nasmyth
platform, only azimuthal motion will change the effective direction of
such a field with respect to the cryostat. It is also worth noting
that the absolute azimuth of the telescope was close to 180$^\circ$
(i.e., nearly due-south). We suspect that this observing
configuration, combined with the large amplitude of the scan, may have
conspired to produce these large signals.

@@

The S2SRO period took place during February and March 2010. For the
purpose of this paper, we will consider data taken as early as 2009
December 3, at which point the instrument was being commissioned in a
similar configuration to that of the S2SRO period. At this time, each
of the 450 and 850\,\micron\ focal planes were populated with single
subarrays, s4a and s8d, respectively. There were two main mechanical
problems to contend with during this period. First, the fridge was not
able to cool the focal plane to the desired base temperature of
... which affected stability, particularly at 850\,\micron. Second,
there were problems with the subarray fabrication process, which led
to large numbers of dead pixels, as well as large-scale variations in
$T_\mathrm{C}$ and the thermal conductivity, $G$. These latter issues
made it difficult to simulataneously place all of the TESs into the
optimal region of the superconducting transition, leading to
significantly poorer sensitivities.

While there were moderate variations from night-to-night during the
observing period, there were typically 600 to 900 working detectors in
each of the s4a and s8d subarrays. Fig.~\ref{fig:sensitivities} shows
representative responsivity maps for the two subarrays from the middle
of S2SRO, based on the response to ramping the bolometer heaters.

@@

The basic method is as follows: (i) a covariance matrix is built up
for all pairs $(i,j)$ of the $N$ bolometer time series,
$\langle\mathbf{b}_i(t),\mathbf{b}_j\rangle$; and (ii) a singular
value decomposition is used to identify a new set of statistically
uncorrelated eigenvectors, $\mathbf{\xi}_i(t)$ (i.e. whose covariance
matrix is diagonal), such that each of the bolometer time series is a
linear combination of the eigenvectors, or \emph{components}
i.e. $\mathbf{b}_i(t) = \bar{\mathbf{\xi}}
\mathbf{\lambda}_i^\mathrm{T}$, where each row of the matrix
$\bar{\mathbf{\xi}}$ is an eigenvector, and
$\mathbf{\lambda}_i^\mathrm{T}$ is a column vector containing the
corresponding eigenvalues. The $\mathbf{\xi}_i$ are normalized by
their \rms, so that the amplitude of each component is stored in the
eigenvalues. In the earlier analyses mentioned, the low-frequency
noise is assumed to be encapsulated in those components with the
largest eigenvalues. Removing the projection of the time series along
those components has been used successfully to reduce $1/f$ noise
while retaining most of the signal in point-like sources. It is also
worth noting that there is an implicit assumption in this analysis
that the noise properties are stationary.

@@

The last two components in Fig.~\ref{fig:pca} have an entirely
different character. The amplitudes of both components are greatest in
low-responsivity regions of the two subarrays (lower-right at
450\,\micron\ and bottom-left at 850\,\micron). Bolometers in these
regions of the subarray are probably only tenuously locked, and are
more sensitive to changes in the loading, possibly resulting in the
step and spikes that are seen in Component 5. Component 6 is clearly
related to Component 5, though with the absence of the steps and
spikes, so that together they seem to represent a continuum of shapes
that are linear combinations of the spikes and steps, with the slower
drift evident only in Component 6.

@@

As described in the introduction, rather than pursue a slow but
potentially more accurate \emph{maximum likelihood} method that would
require the inversion of a complex linear problem, we instead use an
iterative approach that estimates and removes most of the correlated
(and in particular low-frequency) noise sources in parallel with a
simplified map estimation. We do note, however, that recent advances
in direct inversion methods such as SANPIEC \citep{patanchon2008} may
in fact be tenable for \scuba\ data. A future goal of the software
team is to investigate meshing SMURF (perhaps as a cleaning step and
initial condition) for a SANEPIC-like algorithm.

@@

Sudden steps can occur in the time stream data from each bolometer.

\emph{why?}

>>>>

The Submillimetre Common User Bolometer Array 2
\citep[\scuba,][]{2006SPIE.6275E..45H,holland2012} is a new instrument
for the 15-m James Clerk Maxwell Telescope (JCMT) on Mauna Kea,
Hawai'i. The camera simultaneously images the sky in two broad bands
centered over 450 and 850\,\micron, with approximately 7.5 and
14.5\,arcsec full-width at half-maximum (FWHM) point spread functions
(PSFs). The focal planes at each wavelength are populated with 4
rectangular subarrays, consisting of $40 \times 32$ bolometers each,
and together subtend a nearly 7\,arcmin $\times$ 7\,arcmin footprint
on the sky (excluding gaps, the continuous solid angle is about
43\,arcmin$^{2}$ per focal plane).  The instrument is also equipped
with a Fourier transform spectrometer \citep[FTS-2,][]{gom2010}, and a
polarimeter \cite[POL-2,][]{2011ASPC..449...68B}. This paper describes
the properties of \scuba\ data that are relevant to producing maps of
the imaging data, and the Submillimetre User Reduction Facility,
SMURF, a software package for perfoming the reduction written using
the Starlink Software Environment \citep{2009ASPC..411..418J}. The
analysis of FTS-2 and POL-2 data will be described at a later date
once these additional instruments have been commissioned and
characterized. The details of the instrument design, performance, and
its calibration are given in two companion papers: \citet{holland2012}
and \citet{dempsey2012}.

@@

The \scuba\ bolometer temperatures are read-out using superconducting
transition-edge sensors (TESs), with critical temperatures
$T_\mathrm{C}$ approximately 140 and 200\,mK at 850 and 450\,\micron,
respectively. The range about $T_\mathrm{C}$ in which the transition
from a superconductor to a normal material occurs is extremely small
(...typical value...?). The entire \scuba\ focal plane is held at a
\emph{base temperature} somewhat below $T_\mathrm{C}$. The TESs are
then biased up into the transition using two systems. First,
individual pixel heaters (using a current bias to apply the desired
Joule heating) are set to fixed values at the start of an observation
to compensate for most of the loading from the instrument and
atmosphere (in better weather, less compensation from the heaters is
required). Then, an electrical voltage bias is applied. As the
instrument observes, and there are tiny temperature fluctuations due
to astronomical sources and atmosphere, the impedence
changes. Thermoelectrical feedback helps to keep the TESs ``locked''
in the transition as this occurs; an increase in temperature increases
the impedence, and lowers the current and resulting heating
power. Changing magnetic fields resulting from these current changes
are then amplified using chains of superconducting quantum
interference devices (SQUIDs), and the larger output current is
finally digitized. Each detector has its own SQUID for the first-stage
of the amplification, but the remaining stages occur within a common
chain of SQUIDs for each of the 32 columns. All 40 rows are then
read-out in sequence at a row-visit frequency of about 12\,kHz. This
multiplexing scheme greatly reduces the number of readout and bias
lines to each detector, making such large-format bolometer array
designs feasible. It should also be noted that there is an additional
41st row of SQUID readouts that are not connected to TES. These ``dark
SQUIDs'' are used to track non-thermal noise sources that are common
to each column's amplifier chain.

The response of the output digitized current, $I$, with respect to a
change in the power $P$ incident on the bolometer, $dI/dP$ depends on
the efficiency of the thermal absorber, how well the TES is biased
into the transition, and the gain of the SQUID amplifiers. In
practice, the response of a detector can therefore vary somewhat from
observation to observation depending on how the arrays have been
tuned, and the observing conditions \citep[see][]{holland2012}. To
track these changes, at the start of each observation, the current is
ramped through the bolometer heaters such that $dI/dP$ can be
measured. We will refer to the measurements of $dI/dP$ across the
focal plane as the \emph{flatfield} for a given observation.

In general, bolometers have two predominant noise components: (i) the
effectively white thermal noise produced by the phonon and optical
loading on the absorbers (largely independent from bolometer to
bolometer); and (ii) and longer timescale drift resulting from slow
variations in the mean load (e.g. fridge, telescope, or atmospheric
changes), and also the readout electronics (often correlated amongst
many or all of the bolometers). As such, the spectrum corresponding to
a bolometer time-series exhibits a ``$1/f$'' shape below some knee
frequency, $f_\mathrm{knee}$, and is relatively flat at higher
frequencies. Other smaller effects may include line features in the
spectrum (a common culprit is pickup from AC mains, or microphonic
signals induced by mechanical vibrations), and finally a roll-off
towards the Nyquist frequency caused by the anti-aliasing filter that
precedes digitization.

The final conversion of power received by the detector into
astronomical flux units requires two more fundamental corrections: (i)
atmospheric extinction (due to atmospheric water vapour) must be
compensated for as a function of time; and (ii) the corrected power is
related to an astronomical flux density through the effective area of
the detector, the spectral response of the bolometers (with bandpasses
set by optical filters), the spectrum of the source, and the optical
efficiencies of the dish surface and other optical components in the
system -- a \emph{flux conversion factor} (FCF). Extinction is
measured, in most cases, using the JCMT line-of-sight water vapour
radiometer (WVM) which is calibrated using sky dips and the CSO
225\,GHz tipper. The FCF is determined empirically using measurements
of standard sources. For further details on these calibration steps,
see \citet{dempsey2012}.

@@

In order to observe astronomical sources, it is clearly advantageous
that the signals appear in the region of the spectrum that offers the
highest \snr, namely frequencies above $f_\mathrm{knee}$. This effect
is achieved by modulating the signal; the original SCUBA primarily
used the JCMT chopping secondary to place sources into the spectrum at
about $\sim$10\,Hz. The problem with this strategy was that SCUBA was
ultimately limited to recovering information on angular scales smaller
than the chop throw \citep[up to several arcmin,][]{holland1999}.  The
primary approach to map-making with \scuba\ is instead to slew the
entire telescope at such a rate that angular scales of interest are
crossed on time scales shorter than $1/f_\mathrm{knee}$, as with other
earlier ground-based instruments that did not use a chopping secondary
\citep[e.g.,][]{glenn1998,weferling2002,wilson2008,kovacs2008b}. The
desire to scan rapidly has two basic constraints: (i) while the
telescope can physically move up to about 1800 and
3600\,arcsec\,s$^{-1}$ in azimuth and elevation, respectively, the
acceleration limit is about 600\,arcsec\,s$^{-2}$ (in azimuth, larger
in elevation), meaning that significant time is waisted on
``turn-arounds'' at large scan speeds; and (ii) bolometers are
designed with a thermal time constant that balances response with
sensitivity. Taking these constraints into consideration, \scuba\ was
ultimately designed to operate at a maximum scan speed of
600\,arcsec\,sec$^{-1}$. The data acquisition system has a target
sample rate of 200\,Hz (although in practice it is typically in the
range 170--180\,Hz), which at that scan velocity results in a sample
every 3.0\,arcsec, or approximately one third of the 450\,\micron\
diffraction-limited full-width at half-maximum (FWHM) -- a typical
rule-of-thumb for adequately sampling a Gaussian point spread function
(PSF). Similarly, the \scuba\ bolometers are designed to have a time
constant of $\sim$1\,ms in order to avoid distorting PSF-sized sources
at this maximum scan rate.

The scan strategy is designed, primarily, to mitigate the effects of
$1/f$ drift. The simplest scan one might use is a raster
``boustrophedon'' pattern, i.e. a series of parallel passes across the
field. However, such a scan yields extremely poor performace for
structures that are transverse to the scan direction, since a long
time will have ellapsed between each parallel step, and the signal
will therefore be totally dominated by the $1/f$ noise. In addition,
other periodic noise features may appear in the data. With such a
regularly spaced scan strategy, these noise features can easily
conspire with the scan pattern to produce similarly periodic noise
structures in the map, or \emph{scan-synchronous} noise. It has
therefore been recognized that scan strategies that visit each region
of the map on as many different time scales, and scan angles as
possible offer the best protection against large-scale noise
\citep[see discussion in][and references therein]{kovacs2008b}; such
strategies are said to offer good \emph{cross-linking}. As a tradeoff
between truly random scan patterns and feasibility of implementation
taking into consideration both mechanical and scheduling constraints,
\scuba\ presently has three primary scan modes: PONG, which covers
rectangular areas larger than the array footprint in a closed pattern
that ``bounces'' off the edges at 45\,degree angles (and also usually
combined with a rotation through a number of fixed position angles to
create a ``rotating PONG'' with even better cross-linking); Lissajous,
in which sinusoidal functions describe the motion along orthogonal
axes (also to cover rectangular areas -- the PONG and Lissajous modes
are both related to similar scan modes developed at the Caltech
Submillimetre Observatory); and a ``constant velocity daisy'' for
small areas (of order the array footprint), in which the telescope
moves in a circle, whose centre also slowly traces out a circle. For a
more complete description of the \scuba\ observing modes, see
\citet{holland2012}.

@@

Next, we produce power spectral density (PSD) plots for several
representative bolometers in Fig.~\ref{fig:pspec}. To produce this
figure, we follow the convention that the PSD as a function of
frequency, $\mathbf{P}(f)$, is normalized such that the integral over
frequency gives the same variance as the time-series variance across
the full time series. In other words, given a bolometer signal
$\mathbf{b}(t)$,
%
\begin{equation}
\label{eq:psd}
\langle\mathbf{b}^2(t)\rangle = 2 \int_0^{f_\mathrm{N}} \mathbf{P}(f)
df .
\end{equation}
%
where we only integrate over the positive frequencies up to
$f_\mathrm{N}$, the Nyquist frequency, and the factor of 2 accounts
for the (equal) power that appears at negative frequencies in the
discrete Fourier Transform. The units of the PSD written in this form
are then pW$^2$\,Hz$^{-1}$; without the factor of 2 in front the
convention would be to write the units as pW$^2$\,s.

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{pspec.pdf}
\caption{Bolometer power spectral densities. Four representative
  bolometers have been selected from each sub-array, and the
  flat-fielded (but otherwise raw) PSDs are shown as coloured dotted
  lines (the blue signals are for the same time series as those shown
  in Fig.~\ref{fig:bolos_mix}). Horizontal dotted lines at $5 \times
  10^{-8}$ and $5 \times 10^{-9}$\,pW\,Hz$^{-1}$ at 450 and
  850\,\micron, respectively, give the approximate white-noise
  levels. There are broad line features in the PSDs at both
  wavelengths above $\sim$35\,Hz. From $\sim$50\,Hz up to the Nyquist
  frequency, the gradual roll-off of the anti-aliasing filter is also
  evident.  At lower frequencies, the bolometer signals exhibit clear
  $1/f$ knees at approximately 2 and 4\,Hz at 450 and
  850\,\micron. The solid black lines are the PSDs of the common-mode
  signals at each wavelength, and the solid coloured lines show the
  PSDs of the bolometers once the common-mode is removed. These
  residual signals have significantly lower $1/f$ knees, approximately
  0.4 and 2\,Hz at 450 and 850\,\micron. Finally, the dashed black
  lines show the spectral shape produced by a point source, assuming
  that the telescope is slewing at 200\,arcsec\,sec$^{-1}$ (typical
  for S2SRO); for reference, the top horizontal axis shows the
  conversion from frequency to angular scale. We can see that the line
  features may add significant noise on the scale of the PSF. We also
  see that, for this scan velocity, the low-frequency noise begins to
  dominate on scales of approximately 8.3 and 1.7\,arcmin; these
  scales are comparable to the array footprint. .7\,arcmin; these
  scales are comparable to the array footprint. \bf{Per:The array foot
  print can't both be 1.7 and 8.3 arc in unless you used 4 arrays at
  450 and only one at 850?}}
\label{fig:pspec}
\end{figure*}

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded data. At each wavelength, there is a clear
$1/f$ knee at a few Hz, followed by a predominantly white spectrum
punctuated by line features above $\gsim 30$\,Hz, and finally a
roll-off caused by the anti-aliasing filter above $\gsim 70$\,Hz. As
indicated in the previous section, the correlation between the
low-frequency components of the different bolometer signals is
large. The solid black lines in Fig.~\ref{fig:pspec} indicate the PSDs
of the common-modes $\mathbf{c}(t)$ at each wavelength, which
reproduce most of the low-frequency structure, as well as the
high-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high-frequency,
as expected if the bolometers are dominated by un-correlated white
noise. Finally, we note that the amplitudes and slopes of the
$\mathbf{c}(t)$ are similar at both 450 and 850\,\micron. However,
since the 450\,\micron\ white noise is larger than at 850\,\micron\
(shown approximately by the horizontal dotted lines), the $1/f$ knee
occurs at a \emph{lower} frequency at 450\,\micron.

@@

An example of an observation where pickup appears to be significant is
shown in Fig.~\ref{fig:magpickup}. The time-series for two
450\,\micron\ bolometers in the same column (not flatfielded) show
that there is a strong signal with a similar shape, but opposite
signs. This behaviour is seen across the entire array. The dark squid
signal for the same columns exhibits a similar shape and
amplitude. Since the dark squid has no thermal absorber or TES
attached to it, this observed signal is not likely to be optical or
thermal in nature (although there can be some cross-talk with the
bolometers). It is also for this reason that the higher-frequency
noise that appears in the bolometer data (thermal) does not appear in
the dark squid, leading to a significantly higher \snr. Due to the
fact that the sign of the gain in each stage of SQUID amplification is
random (although the combined gain is constrained to be negative), and
since magnetic field pickup is only seen at the input to the second
stage, the pickup can appear with random signs for bolometers along a
a column, giving it a distinct signature from other common signals
that appear with the same sign.

@@

The telescope pointing offsets for this approximately 0.5\,degree
diameter scan are also shown in Fig.~\ref{fig:magpickup}. Since the
phase of the azimuth offsets from the map centre in this scan pattern
slowly drifts with respect to the elevation offsets, it is clear that
the bolometer and dark squid signals are detecting a noise source that
is correlated only with the azimuthal motion and \emph{not} the
elevation. This behaviour would be expected if if there were a strong
magnetic field fixed with respect to the telescope dome (i.e., the
earth's magnetic field). Since \scuba\ is mounted on a Nasmyth
platform, only azimuthal motion will change the effective direction of
such a field with respect to the cryostat. Tests have shown that, as
in this example, large scans in azimuth generically produce pickup. In
contrast, and as expected, changes in elevation results in pickup that
is approximately three orders-of-magnitude smaller.

@@

The S2SRO period took place during February and March 2010. For the
purpose of this paper, we will consider data taken as early as 2009
December 3, at which point the instrument was being commissioned in a
similar configuration to that of the S2SRO period. At this time, each
of the 450 and 850\,\micron\ focal planes were populated with single
subarrays, s4a and s8d respectively. There were three main problems to
contend with during this period. First, the fridge was not able to
cool the focal plane to the desired base temperature (partly due to
the heat conductance of the wires from the arrays during this period)
which affected stability, particularly at 850\,\micron. Second, a
non-sinusoidal, large-scale oscillation in the fridge temperature
could easily be seen as a common signal to all of the detectors with a
period of approximately 25\,s.  Third, there were problems with the
subarray fabrication process which led to large numbers of dead
pixels, as well as large-scale variations in $T_\mathrm{C}$ and the
thermal conductivity, $G$. These latter issues made it difficult to
simulataneously place all of the TESs into the optimal region of the
superconducting transition, leading to significantly poorer
sensitivities. While there were moderate variations from
night-to-night during the observing period, there were typically 600
to 900 working detectors in each of the s4a and s8d subarrays.

Since S2SRO the fridge performance has been greatly improved using NTD
thermometry to servo the temperature to a constant value, and with new
wiring adequate base temperatures can now be achieved. However, the
variability across the focal plane continues to be a problem now that
there is a full complement of subarrays (which still includes the two
subarrays used during S2SRO). Optimal tuning to achieve the greatest
overall array sensitivity is described in \citet{holland2012}.

@@

The basic method is as follows: (i) a covariance matrix is built up
for all pairs $(i,j)$ of the $N$ bolometer time series,
$\langle\mathbf{b}_i(t),\mathbf{b}_j(t)\rangle$; and (ii) a singular
value decomposition is used to identify a new set of statistically
uncorrelated eigenvectors, $\mathbf{\xi}_i(t)$ (i.e. whose covariance
matrix is diagonal), such that each of the bolometer time series is a
linear combination of the eigenvectors, or \emph{components}
i.e. $\mathbf{b}_i(t) = \bar{\mathbf{\xi}}
\mathbf{\lambda}_i^\mathrm{T}$, where each row of the matrix
$\bar{\mathbf{\xi}}$ is an eigenvector, and
$\mathbf{\lambda}_i^\mathrm{T}$ is a column vector containing the
corresponding eigenvalues. The $\mathbf{\xi}_i$ are normalized by
their \rms, so that the amplitude of each component is stored in the
eigenvalues. In the earlier analyses mentioned, the low-frequency
noise is assumed to be encapsulated in those components with the
largest eigenvalues. Removing the projection of the time series along
those components has been used successfully to reduce $1/f$ noise
while retaining most of the signal in point-like sources. It is also
worth noting that there is an implicit assumption in this analysis
that the noise properties are stationary.

@@

The last two components in Fig.~\ref{fig:pca} have an entirely
different character. The amplitudes of both components are greatest in
low-responsivity regions of the two subarrays (lower-right at
450\,\micron\ and bottom-left at 850\,\micron). Bolometers in these
regions of the subarray are probably not optimally biased, and are
more sensitive to changes in the loading, possibly resulting in the
step and spikes that are seen in Component 5. Component 6 is clearly
related to Component 5, though with the absence of the steps and
spikes, so that together they seem to represent a continuum of shapes
that are linear combinations of the spikes and steps, with the slower
drift evident only in Component 6.

@@

As described in the introduction, rather than pursue a slow but
potentially more accurate \emph{maximum likelihood} method that would
require the inversion of a complex linear problem, we instead use an
iterative approach that estimates and removes most of the correlated
(and in particular low-frequency) noise sources in parallel with a
simplified map estimation. We do note, however, that recent advances
in direct inversion methods such as SANEPIC \citep{patanchon2008} may
in fact be tenable for \scuba\ data. A future goal of the software
team is to investigate meshing SMURF (perhaps as a cleaning step and
initial condition) for a SANEPIC-like algorithm.

@@

Sudden steps can occur in the time stream data from each bolometer,
with the most likely cause being wraparound in the first stage squid
amplification, the details of which are given in \citet{holland2012}
\textbf{Per gave a technical explanation, but this is probably best in
another paper?}.
